Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-01-19 manuscript/manuscript.qmd
Edited through Study 2.

We now examine whether advisers take the advisees' expectations into
account. Advisers observe the true performance of the advisee, such that
expectations about performance add no additional information. However,
because participants who score higher are also more likely to pick the
High Performer Group, we use our Default treatment as an instrument.
We show linear probability models for the decision of the Advisers, which is
equal to 1 if they recommend competing against the High Performer Group and
0 if they recommend the Low Performer Group. In Column 1 of @tbl-study1regs,
we show our main specificiation with the experimental assignment for Advisees
(High Performer Default vs. Low Performer Default) and Advisers
(Expectation Shown vs. Expectation Hidden), as well as their interaction.
Because each adviser made ten
recommendations, we cluster standard errors at the adviser level. We find that
displaying expectations of those in the Low Performer Default reduces recommendations
to compete against the Top group by 4 percentage points, or about 10\% relative
to when expectations are hidden. This decrease is marginally smaller advisees
with a default choice to compete against the High Performer Group
(nudged to have high expectations, though not necessarily holding them)
were more likely to receive advice to compete against stronger
competitors when their expectations were shown compared to when they
were hidden. We reported a linear probability model on advising to
compete against the High Performer group for advisers in the
"Expectations" treatment, controlling for the true performance of the
advisee in Column 2 of @tbl-study1regs. We observe a significant main
effect of the default treatment, where a high default increased the
likelihood of receiving flattering advice. It is worth noting that,
since only
`r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100`
% of advisees in the high default group actually stuck with their
default choice, this represents a conservative test of the effect of
high expectations on the likelihood of receiving flattering advice.

```{r}
#| label: fig-study1Advice
#| fig-cap: "Advisees who were nudged to express high expectations were more likely to receive advice to compete against the High Performer Group."

study1_time2_advice %>%
  mutate(Treatment=factor(Treatment, levels=c('High','Low'), labels = c('High default', 'Low default'))) %>%
  group_by(Condition, Treatment) %>%
  summarize(avg_count = mean(Advice), ci = qnorm(0.975)*sd(Advice)/sqrt(dplyr::n())) %>%
  ggplot2::ggplot(ggplot2::aes(x = Condition, y = avg_count, fill = Treatment)) +
  ggplot2::geom_col(position = 'dodge') +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = avg_count - ci, ymax = avg_count + ci), 
                         position = ggplot2::position_dodge(0.9), linewidth = 0.2, width = 0.1) +
  ggtitle("Advice: Meeting Expectations vs. Absent Expectations") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  theme_bw() +
  labs(y = "P(Compete Against High Performer)", x = NULL) +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        axis.text.x = element_text(size = 12),
        legend.position = "none", # No legend
        plot.title = element_text(hjust = 0.5, size = 16)) +
  scale_x_discrete(labels = c("Expectation shown", "Expectation Absent")) +
   geom_text(aes(label=Treatment, y = -0.01), position = position_dodge(0.90))

@@

When the quality of a decision does not depend on someone's
expectations, the adviser's awareness of the advisee's expectations
should not influence the advice given. However, in line with our
argument that advisers take into account the belief utility of the
advisee and prefer to flatter rather than disappoint, we find that
displaying the advisee's expectations does make a difference.
Specifically, advisees who expressed a preference for stronger
competitors were more likely to receive 'flattering' advice, encouraging
them to pursue that option. One possible explanation for this finding is
that, rather than trying to flatter, advisers may simply be telling
advisees what they want to hear. In the next study, we aim to
investigate the mechanism behind this discrepancy in advice-giving when
expectations are revealed, exploring the question: Why do people give
flattering advice?

@@

So far, we have shown that displaying expectations influences the advice
given. Specifically, advisers were more likely to recommend competing
against a group of high performers to advisees with higher performance
expectations compared to those without. We now directly test whether
this effect is driven by interpersonal concerns. In particular, we
explore whether more flattering advice stems from a desire to be liked,
and whether advisees, in turn, like advisers more when they receive
flattering advice.

>>>>

We now examine whether advisers take the advisees' expectations into
account. Advisers observe the true performance of the advisee, such that
expectations about performance add no additional information. However,
because participants who score higher are also more likely to pick the
High Performer Group, we use our Default treatment as an instrument that reflect
higher expectations.
We show linear probability models for the decision of the Advisers, which is
equal to 1 if they recommend competing against the High Performer Group and
0 if they recommend the Low Performer Group. In Column 1 of @tbl-study1regs,
we show our main specificiation with the experimental assignment for Advisees
(High Performer Default vs. Low Performer Default) and Advisers
(Expectation Shown vs. Expectation Hidden), as well as their interaction.
Because each adviser made ten
recommendations, we cluster standard errors at the adviser level. We find that
displaying expectations of those in the Low Performer Default reduces recommendations
to compete against the Top group by 4 percentage points, or about 10\% relative
to when expectations are hidden. This decrease is marginally smaller for advisees
with a default choice to compete against the High Performer Group
(nudged to have high expectations, though not necessarily holding them), and whose
advice did not differ based on whether their expectations were shown or hidden.
We reported a linear probability model on advising to
compete against the High Performer group for advisers in the
"Expectations" treatment, controlling for the true performance of the
advisee in Column 2 of @tbl-study1regs. We observe a significant main
effect of the default treatment, where a high default increased the
likelihood of receiving flattering advice. It is worth noting that,
since only
`r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100`
% of advisees in the high default group actually stuck with their
default choice, this represents a conservative test of the effect of
high expectations on the likelihood of receiving flattering advice.

```{r}
#| label: fig-study1Advice
#| fig-cap: "Advice given in Study 1. Advisees who were nudged to express high expectations were more likely to receive advice to compete against the High Performer Group."

study1_time2_advice %>%
  mutate(Treatment=factor(Treatment, levels=c('High','Low'), labels = c('High default', 'Low default'))) %>%
  group_by(Condition, Treatment) %>%
  summarize(avg_count = mean(Advice), ci = qnorm(0.975)*sd(Advice)/sqrt(dplyr::n())) %>%
  ggplot2::ggplot(ggplot2::aes(x = Condition, y = avg_count, fill = Treatment)) +
  ggplot2::geom_col(position = 'dodge') +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = avg_count - ci, ymax = avg_count + ci), 
                         position = ggplot2::position_dodge(0.9), linewidth = 0.2, width = 0.1) +
  # ggtitle("Advice: Meeting Expectations vs. Absent Expectations") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  scale_y_continuous("P(Compete Against High Performer)",
                     labels = scales::percent) +
  geom_text(aes(x = Condition, y = 0.02,
                label = paste0(round(avg_count*100), "%")),
            position = position_dodge(0.9),
            color = "white") +
  papaja::theme_apa() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        axis.text.x = element_text(size = 12),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", # No legend
        plot.title = element_text(hjust = 0.5, size = 16)) +
  scale_x_discrete(labels = c("Expectation Shown", "Expectation Hidden")) +
   geom_text(aes(label=Treatment, y = -0.01), position = position_dodge(0.90))

@@

When an adviser knows an outcome that an advisee can only estimate, learning
this estimate should not change the advice that is given. However, in line with our
argument that advisers take into account the belief utility of the
advisee and prefer to avoid disappointing them, we find that providing information
about expectations makes a difference.
Specifically, advisees who expressed a preference for stronger
competitors were more likely to receive flattering advice, encouraging
them to pursue that option. This occurs even though advisers are incentivized
to give good advice.

@@

We next examine if advisers believe that flattering advice will improve
how advisees perceive them, and whether this perception is true. We do this
by manipulating the incentives for advisers, who receive a bonus either based
on how they are evaluated by an advisee, or a bonus for the advisee's accuracy.
